Exhaustive dynamic analysis of reachability and stable states of biological regulatory networks modeled in Process Hitting using Answer Set Programming

### 2015-04-28 main.tex
relecture complète

@@

\begin{document}
\bibliographystyle{acmtrans}

@@

\label{firstpage}
%
\begin{abstract}
In this paper, we propose a new logical approach to perform model-checking on a restriction of synchronous automata networks, namely Process Hitting. The Process Hitting framework is notably suitable, but not limited, to model and analyze efficiently biological regulatory networks. 
The originality of our model-checking approach relies in the use of Answer Set Programming to consider exhaustive enumeration of all possible solutions of a property with respect to a given execution length.  Our work here focuses on state reachability properties (which are equivalent to the CTL “$\mathsf{EF}$” operator with paths limited to a given length) on the one hand, on the identification of fixed points on the other hand.
The merits of our methods is illustrated by applying them to
biological examples of various sizes
and comparing the results with some existing approaches.
It turns out that our approach succeeds in processing large models
with a high number of components and interactions.

@@

\section{Preliminary definitions}
\label{defs}
%\input{ASP}
\subsection{Process Hitting}
\input{parts/PH}

\section{Fixed point enumeration}
\label{fixpoint}
\input{parts/fixPoint}

\section{Dynamical analysis}
\label{dynamics}
\input{parts/reachability}

\section{Comparative performance analysis}
\label{comparison}
\input{parts/comparaison}

@@

\section{Conclusion and future directions}
\label{ccl}
\input{parts/cc}

>>>>

% Pretty ref
\usepackage{prettyref}
\newrefformat{def}{Definition~\ref{#1}}
\newrefformat{fig}{Figure~\ref{#1}}
\newrefformat{tab}{Table~\ref{#1}}
%\newrefformat{pro}{Property~\ref{#1}}
%\newrefformat{pps}{Proposition~\ref{#1}}
%\newrefformat{lem}{Lemma~\ref{#1}}
\newrefformat{th}{Theorem~\ref{#1}}
\newrefformat{sec}{Section~\ref{#1}}
%\newrefformat{subsec}{Subsect.~\ref{#1}}
%\newrefformat{suppl}{Appendix~\ref{#1}}
\newrefformat{ex}{Example~\ref{#1}}
%\newrefformat{eq}{Eq.~\eqref{#1}}
\def\pref{\prettyref}

@@

\begin{document}
%\bibliographystyle{acmtrans}

@@

\label{firstpage}
%
\begin{abstract}
In this paper, we propose a new logical approach to perform model-checking on a restriction of synchronous automata networks, namely Process Hitting. The Process Hitting framework is notably suitable, but not limited, to model and analyze efficiently biological regulatory networks. 
The originality of our model-checking approach relies in the use of Answer Set Programming to consider exhaustive enumeration of all possible solutions of a property with respect to a given execution length.  Our work here focuses on state reachability properties (which are equivalent to the CTL ``$\mathsf{EF}$'' operator with paths limited to a given length) on the one hand, on the identification of fixed points on the other hand.
The merits of our methods is illustrated by applying them to
biological examples of various sizes
and comparing the results with some existing approaches.
It turns out that our approach succeeds in processing large models
with a high number of components and interactions.

@@

\section{Preliminary definitions}
\label{sec:defs}
%\input{parts/ASP}
\subsection{Process Hitting}
\input{parts/PH}

\section{Fixed point enumeration}
\label{sec:fixpoint}
\input{parts/fixPoint}

\section{Dynamical analysis}
\label{sec:dynamics}
\input{parts/reachability}

\section{Comparative performance analysis}
\label{sec:comparison}
\input{parts/comparaison}

@@

\section{Conclusion and future directions}
\label{sec:ccl}
\input{parts/cc}

### 2015-04-28 parts/ASP.tex
changement pour ICLP-15
relecture complète

In order to develop these methods we have used two main frameworks. The first is both a programming paradigm and a declarative programming language: the Answer Set Programming (ASP).
The second framework is the Process Hitting (PH) a previously introduced formalism for the representation of biological regulatory networks.

\subsection{Answer Set Programming }

\emph{Answer Set Programming} (ASP) is a declarative programming paradigm with semantics known as the \emph{semantics of answer sets}.
This paradigm allows the programmer to specify what the problem to be solved is, and not how to solve it.
ASP programs are written in AnsProlog* \cite{sureshkumar2006ansprolog} (short for “Answer Set Programming in Logic”).
These programs are composed of a set of \emph{facts} and a set of \emph{rules}, as defined below, from which other facts can be derived.
A consistent set of facts that can be derived from a program using the rules is called \emph{answer set} for the program.
The sets of possible responses to an AnsProlog* program are calculated with a program called a solver.

\subsubsection{Syntax and semantics of ASP}
\label{sectionSyntaxeASP}
\begin{enumerate}

@@

According to \cite{baral2003knowledge}, the alphabet of ASP is composed of seven classes of symbols:
\begin{itemize}
  \item \emph{constants}, symbols of \emph{functions} and symbols of \emph{predicates} are composed of letters and start with a lowercase letter,
  \item \emph{variables} follow the same rule but start with a capital letter,
  \item \emph{connectors} that are “$\leftarrow$”, “$\textbf{not}$” and “$,$”,
  \item \emph{punctuation} that can be “(”, “)” and “.”,
  \item the special symbols “$\bot$” and “$\top$”.
\end{itemize}

The notion of predicate allows to qualify constants in this grammar.
For example, the property “\textit{Tweety is a bird}”
can be expressed by the predicate name \textit{bird}
around the constant \textit{tweety} into the following literal:
\textit{bird(tweety)}.
Such literal can them be used as premise or conclusion in a rule as defined below.

\item \textbf{The rules:}
A rule has the following form:
\begin{equation} \label{eq1ASP}
 head \leftarrow body.
\end{equation}
which is equivalent to:
\begin{equation} \label{eq2ASP}
l_{0} \leftarrow l_{1},...,l_{m}, \ASPnot l_{m+1},..., \ASPnot l_{n}.
\end{equation}
where $l_{i}$ are literals for all $i \in \llbracket 1 ; n \rrbracket$, and $0 \leq m \leq n$.
This rule states:
\[
\left.
    \begin{array}{rcl}
        \text{If } & l_{1},...,l_{m}  & \text{ \textbf{can} be proved true}\\
        \text{and } & l_{m+1},...,l_{n} & \text{ \textbf{cannot} be proved true}
    \end{array}
\right \} \text{Then} ~ l_{0} ~ \text{is } \textbf{true}.
\]

Therefore, “$head$” is also called the \emph{conclusion} of the rule,
and “$body$” is its \emph{premise}.
% An \emph{answer set} of a program (that is, a set of rules)
% is a minimal set of literals that satisfies all the rules.

@@

\item A \emph{constraint} is a rule where the head equals $\bot$.
  In this case, the head is often not represented and the constraint will be written generally as:
\begin{equation} 
  \leftarrow l_{1},...,l_{m}, \ASPnot l_{m+1},..., \ASPnot l_{n}.
  \label{eq6ASP}
\end{equation}
A set of literals $X$ \emph{violates} the constraint \eqref{eq6ASP} if $\{l_{1},...,l_{m}\} \subseteq X$ and $\{l_{m+1},...,l_{n}\} \nsubseteq X$.
Thus, $X$ cannot be an answer set of any program containing this constraint.
% If a program $\pi$ contains such a constraint $r$, then $X$ is an answer set of $\pi$ if and only if $X$ is an \textit{answer set} of $\pi \setminus \{r\}$ and $X$ does not violate $r$.

@@

\subsubsection{Modeling and solving of a problem with ASP}

The construction of models is one of the fundamental components of the scientific process.
It concerns all systems we seek to control.
A model has two main features \cite{Glimpse}:
it is a simplification of a given system
and it allows an action on the system like resolving problems, searching for a path, identifying states and verifying properties...
This concept of model can address another angle issues related to the representation process.
the programs written in ASP follow the strategy of “generate and test”.
Indeed, the modeling process can be regarded as a special form of representation:
\begin{itemize}
  \item First, the system should be presented with facts.
  \item Second, properties of the model can be explained with rules.
  \item Third, all candidate answers can be generated, usually using cardinalities.
  \item Finally, constraints allow to filter the answers and only the sets of predicates satisfying all these conditions will be returned which and considered as \emph{answer sets}.
\end{itemize}

@@

For the work presented in this paper, we worked with \textsc{Clingo} (version 3) which is a combination of the grounder \textsc{Gringo} and the solver \textsc{Clasp}.

>>>>

\section*{Answer Set Programming}

We develop in this appendix the syntax and semantics of \emph{Answer Set Programming} (ASP) that is used in our work to solve declarative programs.
% 
% In order to develop these methods we have used two main frameworks. The first is both a programming paradigm and a declarative programming language: the Answer Set Programming (ASP).
% The second framework is the Process Hitting (PH) a previously introduced formalism for the representation of biological regulatory networks.
%
%\subsection{Answer Set Programming }
%
ASP is a declarative programming paradigm with semantics known as the \emph{semantics of answer sets}.
This paradigm allows the programmer to specify what the problem to be solved is, and not how to solve it.
ASP programs are written in AnsProlog* \cite{sureshkumar2006ansprolog} (short for ``Answer Set Programming in Logic'').
These programs are composed of a set of \emph{facts} and a set of \emph{rules}, as defined below, from which other facts can be derived.
A consistent set of facts that can be derived from a program using the rules is called \emph{answer set} for the program.
The sets of possible responses to an AnsProlog* program are calculated with a program called \emph{solver}.

\subsection*{Syntax and semantics of ASP}
\label{sectionSyntaxeASP}
\begin{enumerate}

@@

According to \cite{baral2003knowledge}, the alphabet of ASP is composed of seven classes of symbols:
\begin{itemize}
  \item \emph{constants}, symbols of \emph{functions} and symbols of \emph{predicates} are composed of letters and start with a lowercase letter,
  \item \emph{variables} follow the same rule but start with a capital letter,
  \item \emph{connectors} that are ``$\leftarrow$'', ``$\textbf{not}$'' and ``$,$'',
  \item \emph{punctuation} that can be ``('', ``)'' and ``.'',
  \item the special symbols ``$\bot$'' and ``$\top$''.
\end{itemize}

The notion of predicate allows to qualify constants in this grammar.
For example, the property ``\textit{Tweety is a bird}''
can be expressed by the predicate name \textit{bird}
around the constant \textit{tweety} into the following literal:
\textit{bird(tweety)}.
Such literal can then be used as premise or conclusion in a rule as defined below.

\item \textbf{The rules}

A rule has the following form:
\begin{equation} \label{eq1ASP}
 head \leftarrow body.
\end{equation}
which is equivalent to:
\begin{equation} \label{eq2ASP}
l_{0} \leftarrow l_{1},...,l_{m}, \ASPnot l_{m+1},..., \ASPnot l_{n}.
\end{equation}
where $l_{i}$ are literals for all $i \in \llbracket 1 ; n \rrbracket$, and $0 \leq m \leq n$.
This rule states:
\[
\left.
    \begin{array}{rcl}
        \text{if } & l_{1},...,l_{m}  & \text{ \textbf{can} be proven true}\\
        \text{and } & l_{m+1},...,l_{n} & \text{ \textbf{cannot} be proven true}
    \end{array}
\right \} \text{ then} ~ l_{0} ~ \text{is } \textbf{true}.
\]

Therefore, ``$head$'' is also called the \emph{conclusion} of the rule,
and ``$body$'' is its \emph{premise}.
% An \emph{answer set} of a program (that is, a set of rules)
% is a minimal set of literals that satisfies all the rules.

@@

\item A \emph{constraint} is a rule where the head equals $\bot$.
  In this case, the head is often not represented and the constraint will be written generally as:
\begin{equation} 
  \leftarrow l_{1},...,l_{m}, \ASPnot l_{m+1},..., \ASPnot l_{n}.
  \label{eq6ASP}
\end{equation}
A set of literals $X$ \emph{violates} a constraint if
its body is true, and thus cannot be an answer set
(as $\bot$ cannot be part of an answer set).
%$\{l_{1},...,l_{m}\} \subseteq X$ and $\{l_{m+1},...,l_{n}\} \nsubseteq X$.
%Thus, $X$ cannot be an answer set of any program containing this constraint.
% If a program $\pi$ contains such a constraint $r$, then $X$ is an answer set of $\pi$ if and only if $X$ is an \textit{answer set} of $\pi \setminus \{r\}$ and $X$ does not violate $r$.

@@

\subsection*{Modeling and solving of a problem with ASP}

The construction of models is one of the fundamental components of the scientific process.
It concerns all systems we seek to control.
A model has two main features \cite{Glimpse}:
it is a simplification of a given system
and it allows an action on the system like resolving problems, searching for a path, identifying states and verifying properties...
This concept of model can address another angle issues related to the representation process.
The programs written in ASP follow the strategy of ``generate and test''.
Indeed, the modeling process can be regarded as a special form of representation:
\begin{itemize}
  \item First, the system should be presented with facts.
  \item Second, properties of the model can be explained with rules.
  \item Third, all candidate answers can be generated, usually using cardinalities.
  \item Finally, constraints allow to filter the answers and only the sets of predicates satisfying all these conditions will be returned which and considered as \emph{answer sets}.
\end{itemize}

@@

For the work presented in this paper, we have worked with \textsc{Clingo} (version 3) which is a combination of the grounder \textsc{Gringo} and the solver \textsc{Clasp}.

### 2015-04-28 parts/PH.tex
relecture complète

The concurrent interactions between processes are defined by a set of \emph{actions}.
Each action is responsible for the replacement of one process by another of the same sort
conditioned by the presence of at most one other process in the current state.
An action is denoted by $\PHfrappe{a_i}{b_j}{b_k}$, which is read as
“$a_i$ \emph{hits} $b_j$ to make it \emph{bounce} to $b_k$”,
where $a_i$, $b_j$, $b_k$ are processes of sorts $a$ and $b$,
called respectively \emph{hitter}, \emph{target} and
\emph{bounce} of the action.
We also call a \emph{self-hit} any action whose hitter and target sorts are the same,
that is, of the form: $\PHfrappe{a_i}{a_i}{a_k}$.

@@

\begin{example*}
\pref{fig:ph} represents a $\PH$ $(\PHs,\PHl,\PHa)$ with three sorts
($\PHs = \{a, b, c\}$) and:
$\PHl_a = \{a_0, a_1\}$,
$\PHl_b = \{b_0, b_1\}$,
$\PHl_z = \{z_0, z_1, z_2\}$.
\begin{figure}[ht]
\centering
\begin{tikzpicture}%[font=\scriptsize]
%\path[use as bounding box] (0,-1) rectangle (4,4);

@@

\path[bounce,bend left]
\TBounce{a_0}{}{a_1}{.south}
;
\TState{a_0,b_0,z_1}
\end{tikzpicture}
\caption{\label{fig:ph} 
A PH model example with three sorts: $a$, $b$ and $z$ ($a$ is either at level 0 or 1, $b$ at either level 0 or 1 and $z$ at either level 0, 1 or 2). Boxes represent the \emph{sorts} (network components), circles represent the \emph{processes} (component levels), and the 5 \emph{actions} that model the dynamic behavior are depicted by pairs of arrows in solid and dotted lines. The grayed processes stand for the possible initial state: $\PHstate{a_1, b_0, z_1}$.
}
\end{figure}
\end{example*}
A state of the networks is a set of active processes containing a single process of each sort.
The active process of a given sort $a \in \PHs$ in a state $s \in \PHl$
is noted $\PHget{s}{a}$.
For any given process $a_i$ we also note: $a_i \in s$ if and only if $\PHget{s}{a} = a_i$.

@@

The PH was chosen for several reasons.
First, it is a general framework that,
although it was mainly used for biological networks,
allows to represent any kind of dynamical model,
and converters to several other representations are available (see section~\ref{comparison}).
Although an efficient dynamical analysis already exists for this framework,
based on an approximation of the dynamics,
it is interesting to identify its limits
and compare them to the approached we present later in this paper.
Finally, the particular form of the actions in a PH model allow
to easily represent them in ASP,
with one fact per action, as described in the next section.
Other representations may have required supplementary complexity;
for instance, a labeling would be required
if actions could be triggered by a variable number of processes.

@@

The notion of \emph{fixed point}, also called \emph{stable state},
is given in definition~\ref{def:fixpoint}.
A fixed point is a state which has no successor.
Such states have a particular interest as they denote states in which the model
stays indefinitely,
and the existence of several of these states denotes a switch in the dynamics~\cite{wuensche1998genomic}.

@@

A finer and more interesting dynamical property consists in
the notion of \emph{reachability property}.
Such a property, defined in definition~\ref{def:reachability},
states that starting from a given initial state, it is possible
to reach a given goal, that is, a state that contains a process
or a set of processes.
Checking such a dynamical property is considered difficult
as in usual model-checking techniques,
it is required to build (a part of) the state graph,
which has an exponential complexity.

@@

\begin{definition}[Reachability property]
\label{def:reachability}
  If $s \in \PHl$ is a state and $A \subseteq \Proc$ is a set of processes,
  we denote by $\mathcal{P}(s, A)$ the following \emph{reachability property}:
  \[\exists? \delta \in \Sce(s), \forall a_i \in A, \PHget{(s \play \delta)}{a} = a_i
    \enspace.\]
\end{definition}

The rest of the paper focuses on the resolution of the previous issues
with the use of ASP.
The enumeration of all fixed points of a PH model will be tackled in
section~\ref{fixpoint}
and the verification of a reachability property will be the subject of
section~\ref{dynamics}.

>>>>

The concurrent interactions between processes are defined by a set of \emph{actions}.
Each action is responsible for the replacement of one process by another of the same sort
conditioned by the presence of at most one other process in the current state.
An action is denoted by $\PHfrappe{a_i}{b_j}{b_k}$, which is read as
``$a_i$ \emph{hits} $b_j$ to make it \emph{bounce} to $b_k$'',
where $a_i$, $b_j$, $b_k$ are processes of sorts $a$ and $b$,
called respectively \emph{hitter}, \emph{target} and
\emph{bounce} of the action.
We also call a \emph{self-hit} any action whose hitter and target sorts are the same,
that is, of the form: $\PHfrappe{a_i}{a_i}{a_k}$.

@@

\begin{example}
\pref{fig:ph} represents a PH $(\PHs,\PHl,\PHa)$ with three sorts
($\PHs = \{a, b, c\}$) and:
$\PHl_a = \{a_0, a_1\}$,
$\PHl_b = \{b_0, b_1\}$,
$\PHl_z = \{z_0, z_1, z_2\}$.
\begin{figure}[ht]
\centering
\begin{tikzpicture}%[font=\scriptsize]
%\path[use as bounding box] (0,-1) rectangle (4,4);

@@

\path[bounce,bend left]
\TBounce{a_0}{}{a_1}{.south}
;
\TState{a_0,b_0,z_1}
\end{tikzpicture}
\caption{\label{fig:ph} 
A PH model example with three sorts: $a$, $b$ and $z$ ($a$ is either at level 0 or 1, $b$ at either level 0 or 1 and $z$ at either level 0, 1 or 2). Boxes represent the sorts (network components), circles represent the processes (component levels), and the 5 actions that model the dynamic behavior are depicted by pairs of arrows in solid and dotted lines. The grayed processes stand for the possible initial state $\PHstate{a_1, b_0, z_1}$.
}
\end{figure}
\end{example}
A state of the networks is a set of active processes containing a single process of each sort.
The active process of a given sort $a \in \PHs$ in a state $s \in \PHl$
is noted $\PHget{s}{a}$.
For any given process $a_i$ we also note: $a_i \in s$ if and only if $\PHget{s}{a} = a_i$.

@@

The PH was chosen for several reasons.
First, it is a general framework that,
although it was mainly used for biological networks,
allows to represent any kind of dynamical model,
and converters to several other representations are available (see \pref{sec:comparison}).
Although an efficient dynamical analysis already exists for this framework,
based on an approximation of the dynamics,
it is interesting to identify its limits
and compare them to the approach we present later in this paper.
Finally, the particular form of the actions in a PH model allow
to easily represent them in ASP,
with one fact per action, as described in the next section.
Other representations may have required supplementary complexity;
for instance, a labeling would be required
if actions could be triggered by a variable number of processes.

@@

The notion of \emph{fixed point}, also called \emph{stable state},
is given in \pref{def:fixpoint}.
A fixed point is a state which has no successor.
Such states have a particular interest as they denote states in which the model
stays indefinitely,
and the existence of several of these states denotes a switch in the dynamics~\cite{wuensche1998genomic}.

@@

A finer and more interesting dynamical property consists in
the notion of \emph{reachability}.
Such a property, defined in \pref{def:reachability},
states that starting from a given initial state, it is possible
to reach a given goal, that is, a state that contains a process
or a set of processes.
Checking such a dynamical property is considered difficult
as in usual model-checking techniques,
it is required to build (a part of) the state graph,
which has an exponential complexity.

@@

\begin{definition}[Reachability property]
\label{def:reachability}
  If $s \in \PHl$ is a state and $A \subseteq \Proc$ is a set of processes,
  we denote by $\mathcal{P}(s, A)$ the following \emph{reachability property}:
  \[\mathcal{P}(s, A) \equiv \exists \delta \in \Sce(s), \forall a_i \in A, \PHget{(s \play \delta)}{a} = a_i
    \enspace.\]
\end{definition}

The rest of the paper focuses on the resolution of the previous issues
with the use of ASP.
The enumeration of all fixed points of a PH model will be tackled in
\pref{sec:fixpoint}
and the verification of a reachability property will be the subject of
\pref{sec:dynamics}.
The semantics of ASP is given in Appendix.
